|

### 2024-12-04 manuscript/paper.Rmd
Added continuous total fights and penality minutes.  Also included an indicator for North America vs Other.

---
title: |
  |  \large Enforcers are fine, probably.  
author: | 
  | \large Zach Binney \vspace{-1.1mm}
  | \normalsize Epi \vspace{-1mm}
  | \normalsize Oxford something at Emory \vspace{-1mm}
  | \normalsize Georgia? \vspace{-1mm}
  | \normalsize [`email`](mailto:ypu@something.edu) \vspace{-1mm}
  |
  | \large Brian M. Mills \vspace{-1.1mm}
  | \normalsize College of Education \vspace{-1mm}
  | \normalsize University of Texas at Austin \vspace{-1mm}
  | \normalsize Austin, TX 78712 \vspace{-1mm}
  | \normalsize [`brian.mills@austin.utexas.edu`](mailto:brian.mills@austin.utexas.edu) \vspace{-1mm}
  |
  | \large Leigha DeRango \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`lderango@luc.edu`](mailto:lderango@luc.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
  | The abstract goes here. \vspace{2mm}
  | Keywords: survival analysis, hockey
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

>>>>

---
title: |
  |  \large Enforcers are fine, probably.  
author: | 
  | \large Zach Binney \vspace{-1.1mm}
  | \normalsize Epi \vspace{-1mm}
  | \normalsize Oxford something at Emory \vspace{-1mm}
  | \normalsize Georgia? \vspace{-1mm}
  | \normalsize [`email`](mailto:ypu@something.edu) \vspace{-1mm}
  |
  | \large Brian M. Mills \vspace{-1.1mm}
  | \normalsize College of Education \vspace{-1mm}
  | \normalsize University of Texas at Austin \vspace{-1mm}
  | \normalsize Austin, TX 78712 \vspace{-1mm}
  | \normalsize [`brian.mills@austin.utexas.edu`](mailto:brian.mills@austin.utexas.edu) \vspace{-1mm}
  |
  | \large Leigha DeRango \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`lderango@luc.edu`](mailto:lderango@luc.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
 |Fighting and Penalty Minutes Associated With Long-term Mortality Among National Hockey League Players, 1967 to 2022
 | A recent study looked at the relationship between long-term mortality and being an enforcer in the National Hockey League.  Specifically, that  study presented a matched cohort study that found that NHL enforcers (defined based on number of career fights or average career penalty minutes) died an everage of 10 years earlier
  \vspace{2mm}
  | Keywords: survival analysis, hockey
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

## Continuous predictors for fights and penalty minutes
```{r}
dat <- dat %>% mutate(northamerica = ifelse(Ctry %in% c("USA","CAN"),"North America","Other"))

ggplot(aes(x = northamerica, y = PIM.GP_numeric), data = dat) + geom_boxplot()

ggplot(aes(x = northamerica, y = Total.Fights), data = dat) + geom_boxplot()

ggplot(aes(color = northamerica,x = PIM.GP_numeric, y = Total.Fights), data = dat) + geom_point() + facet_grid(~northamerica)
```

```{r fig:estsurv, echo = FALSE, fig.cap = "Estimated Survival curve for EP for defenders.  All other variables are fixed at their mean level (GP = 285.8, Ht = 72.7in, Wt = 197.03, and Year of birth 1974.5."}
dat <- dat %>% mutate(northamerica = ifelse(Ctry %in% c("USA","CAN"),"North America","Other"))
coxphmod_PIM.GP_numeric <- coxph(Surv ~ PIM.GP_numeric + GP.x + Ht + Wt + DOB.Year + Pos2, data = dat)
coxphmod_PIM.GP_numeric
```

Mike Lopez suggested looking at North America vs playuers from outside of North America.  
Controlling for North America vs Other penalty minutes, penalty minutes per game is significant wuth a hazard ratio of 1.1880411. 
```{r}
coxphmod_PIM.GP_numeric <- coxph(Surv ~ PIM.GP_numeric + northamerica + GP.x + Ht + Wt + DOB.Year + Pos2, data = dat)
coxphmod_PIM.GP_numeric
```

```{r}
# summary(coxphmod_EP)
# library(xtable)
# xtable(summary(coxphmod_EP)$coef[,-4],digits = 5)
# cox.zph(coxphmod_EP)

newdataEP <-
  with(dat, data.frame(PIM.GP_numeric = c(0.25,.5,3),
                       northamerica = rep("North America",3),
                       GP.x = rep(mean(GP.x), 3),
                       Ht = rep(mean(Ht), 3),
                       Wt = rep(mean(Wt), 3),
                       DOB.Year = rep(mean(DOB.Year), 3),
                       Pos2 = rep("D", 3))
                       )
#plot(survfit(coxphmod_EP, newdata = newdataEP), conf.int = TRUE, lty = c(1,2))

nnn <- nrow(survfit(coxphmod_PIM.GP_numeric, newdata = newdataEP)$surv)
ggdat <- data.frame(PIM.GP_numeric = c(rep(c(.5,3),each = nnn)),
  surv = c(
  survfit(coxphmod_PIM.GP_numeric, newdata = newdataEP)$surv[, 1],
  survfit(coxphmod_PIM.GP_numeric, newdata = newdataEP)$surv[, 2]
),
  lower = c(
  survfit(coxphmod_PIM.GP_numeric, newdata = newdataEP)$lower[, 1],
  survfit(coxphmod_PIM.GP_numeric, newdata = newdataEP)$lower[, 2]
),
  upper = c(
  survfit(coxphmod_PIM.GP_numeric, newdata = newdataEP)$upper[, 1],
  survfit(coxphmod_PIM.GP_numeric, newdata = newdataEP)$upper[, 2]
),
time = rep(survfit(coxphmod_PIM.GP_numeric, newdata = newdataEP)$time,2)
)
ggplot(aes(x = time,y = surv, color = as.factor(PIM.GP_numeric)), data = ggdat) + geom_path() + geom_path(aes(x = time,y = lower, color = as.factor(PIM.GP_numeric)), lty = 3) + geom_path(aes(x = time,y = upper, color = as.factor(PIM.GP_numeric)), lty = 3) 
```

```{r fig:estsurv, echo = FALSE, fig.cap = "Estimated Survival curve for EF for defenders.  All other variables are fixed at their mean level (GP = 285.8, Ht = 72.7in, Wt = 197.03, and Year of birth 1974.5."}
coxphmod_Total.Fights <- coxph(Surv ~ Total.Fights + northamerica + GP.x + Ht + Wt + DOB.Year + Pos2, data = dat)
# summary(coxphmod_EP)
# library(xtable)
# xtable(summary(coxphmod_EP)$coef[,-4],digits = 5)
# cox.zph(coxphmod_EP)

newdataEF <-
  with(dat, data.frame(Total.Fights = c(0,1,7),
                       GP.x = rep(mean(GP.x), 3),
                       Ht = rep(mean(Ht), 3),
                       Wt = rep(mean(Wt), 3),
                       DOB.Year = rep(mean(DOB.Year), 3),
                       Pos2 = rep("D", 3))
                       )
#plot(survfit(coxphmod_EP, newdata = newdataEP), conf.int = TRUE, lty = c(1,2))

nnn <- nrow(survfit(coxphmod_Total.Fights, newdata = newdataEF)$surv)
ggdat <- data.frame(Total.Fights = c(rep(c(0,5,50),each = nnn)),
  surv = c(
  survfit(coxphmod_Total.Fights, newdata = newdataEF)$surv[, 1],
  survfit(coxphmod_Total.Fights, newdata = newdataEF)$surv[, 2],
  survfit(coxphmod_Total.Fights, newdata = newdataEF)$surv[, 3]
),
  lower = c(
  survfit(coxphmod_Total.Fights, newdata = newdataEF)$lower[, 1],
  survfit(coxphmod_Total.Fights, newdata = newdataEF)$lower[, 2],
  survfit(coxphmod_Total.Fights, newdata = newdataEF)$lower[, 3]
),
  upper = c(
  survfit(coxphmod_Total.Fights, newdata = newdataEF)$upper[, 1],
  survfit(coxphmod_Total.Fights, newdata = newdataEF)$upper[, 2],
  survfit(coxphmod_Total.Fights, newdata = newdataEF)$upper[, 3]
),
time = rep(survfit(coxphmod_Total.Fights, newdata = newdataEF)$time,3)
)
ggplot(aes(x = time,y = surv, color = as.factor(Total.Fights)), data = ggdat) + geom_path() + geom_path(aes(x = time,y = lower, color = as.factor(Total.Fights)), lty = 3) + geom_path(aes(x = time,y = upper, color = as.factor(Total.Fights)), lty = 3) 
```

## Birth country
```{r}
epcountry <- dat %>% mutate(northamerica = ifelse(Ctry %in% c("USA","CAN"),"North America","Other")) %>% group_by(EP, northamerica) %>% summarize(n = n()) %>% pivot_wider(names_from = northamerica, values_from = n) %>% mutate(p = `North America`/(`North America`+ Other))

#prop.test(epcountry$`North America`,epcountry$Other+epcountry$`North America`)

#Enforcer by country
efcountry <- dat %>% mutate(northamerica = ifelse(Ctry %in% c("USA","CAN"),"North America","Other")) %>% group_by(EF, northamerica) %>% summarize(n = n()) %>% pivot_wider(names_from = northamerica, values_from = n)  %>% mutate(p = `North America`/(`North America`+ Other))
```
